{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-03 thesis/education.tex
assorted minor revs, intro, notes on reading this, etc
use cleveref

While the previous chapter demonstrated Landslide's bug-finding power
compared to prior MC techniques in a controlled environment,
whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
And while my M.S. thesis~\cite{landslide} showed that students
could annotate P3 Pebbles kernels and thence use Landslide to debug them,
the annotations alone required 2 hours of effort on average per user,
meaning the only students who could benefit were the ones already succeeding enough to have such free time.
Since then, I have extended Landslide with a fully-automatic instrumentation process
for Pebbles thread libraries (P2s) (\sect{\ref{sec:education-pebbles-instrumentation}})
and Pintos kernels (\sect{\ref{sec:education-pintos-instrumentation}})
to improve its accessibility.

I have run several user studies in the Operating Systems classes
at Carnegie Mellon University (CMU), University of Chicago (U. Chicago),
and \revisionminor{The Pennsylvania} State University (PSU),
wherein students get to use Landslide to find and diagnose their own bugs during the semester.
At CMU, I analyzed logs and code snapshots taken as students used Landslide during P2
(\sect{\ref{sec:education-eval-bugfinding}}),
as well as the grades ultimately assigned after students who either did or didn't use Landslide submitted their projects
(\sect{\ref{sec:education-eval-grades}}).
At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s
(\sect{\ref{sec:education-eval-survey}}).
At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
then \revisionminor{they} returned any resulting bug reports to students
(\sect{\ref{sec:education-eval-bugfinding}})
and likewise surveyed them on the quality of the diagnostic output
(\sect{\ref{sec:education-eval-survey}}).

@@

As described in \sect{\ref{sec:landslide-setup}},
all setup from the user's point of view is handled through the {\tt p2-setup.sh} script.%
\footnote{PSU's version is called {\tt psu-setup.sh};
in this section {\tt p2-setup.sh} refers to both unless otherwise noted.
}
It, its helper scripts (\sect{\ref{sec:landslide-glue}}),
and the {\tt landslide} script itself contain several checks to prevent
studence
from accidentally misusing Landslide in ways that could produce mysterious crashes, false bug reports, and so on
(the need for each \revisionminor{check}, as the reader might imagine, discovered through bitter experience).
These include:
\begin{itemize}
	\item {\tt p2-setup.sh} checks if the directory argument correctly points at the top-level P2 basecode directory
		rather than any subdirectories such as {\tt user/libthread/}.
	\item {\tt check-need-p2-setup-again.sh} checks if any source files in the original P2 source directory
		(the argument supplied to {\tt p2-setup.sh}) \revisionminor{have been updated},
		in case the student hoped to fix some bug and verify their fix but forgot to re-run the setup script.
	\item {\tt landslide} checks the supplied test name matches one of the endorsed Landslide-friendly tests
		(students love trying to run Landslide with {\tt racer}, {\tt largetest},
		or even the string {\tt OPTIONS}).
	\item {\tt landslide} checks if any other instance of itself is simultaneously running in the same directory,
		and if so, refuses to do so and advises the student
		to {\tt git clone} the repository afresh for simultaneous use.%
		\footnote{This is ironically implemented with a non-atomic lock file
		and should really be using {\tt flock} instead.
		}
\end{itemize}
\vspace{1em}

\noindent Landslide also includes several P2-specific instrumentations and features to cope with various student irregularities:
\begin{itemize}
	\item Quicksand emits different combinations of {\tt within\_function}/{\tt without\_function} directives
		for Landslide depending on the name of the test.
		For example, for {\tt paradise\_\allowbreak{}lost},
		\revisionminor{designed to test student semaphore code,}
		Landslide will not preempt in a function named {\tt critical\_section()},
		which the test case uses to protect an internal counter used to detect the bug;
		and it will not preempt in any of the {\tt thr\_*()} thread library API functions
		for tests intended to target just the concurrency primitives.
		In future work this could be improved as annotations to be placed inside the test case code itself.
	\item Landslide finds ad-hoc synchronization patterns,
		such as {\tt while (!flag) yield()} or {\tt while (xchg(...)) continue},
		which students often open-code rather than using the prescribed synchronization API,
		and treats them as synchronization points as described in \sect{\ref{sec:landslide-blocking}}.
	\item Landslide finds ``too suspicious'' spinwait-loops in mutex implementations
		which are neither yield- nor {\tt xchg}-loops (as described above),
		which would ordinarily be classified as infinite loop bugs,
		and reports them with a suggestive message ({\tt undesirable\_\allowbreak{}loop\_html()} in {\tt landslide.c})
		referring the student to the appropriate lecture material
		\cite{synchronization-2}.
	\item The {\tt landslide} wrapper script logs the time and command-line options of invocation
		and captures a snapshot of the student code and results of the test and saves them to AFS
		\revisionminor{(CMU's network file system)~\cite{afs}}
		after each run.
		% HURDLE_VIOLATION -- no, this is for p3 (context switcher) features only
\end{itemize}

@@

Landslide ships with several ``approved'' test cases,
i.e., programs copied from, derived from, or at least vaguely resembling
the tests distributed with P2,
which I curated to produce concurrent behaviour suitable for stateless model checking.
Some tests are crafted to target specific bugs which,
from personal experience as a TA, are common in many student submissions;
others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
Many use some of the features/annotations described in \sect{\ref{sec:landslide-testcases}}.

@@

The following tests were released to CMU students:
\begin{itemize}
	\item {\tt broadcast\_test}:
		Tests the {\tt cond\_broadcast()} signalling path with a single waiter.
	\item {\tt mutex\_test}:
		Tests student mutexes under 2 threads with 2 iterations
		(the \revisionminor{second} iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
		This test uses the {\tt TESTING\_MUTEXES}
		described in \sect{\ref{sec:landslide-staticconfig}}
		to enable data-race preemption points within the mutex implementation.
	\item {\tt paradise\_lost}:
		Written for the sake of the Landslide lecture demo
		(\sect{\ref{sec:education-pebbles-recruiting}}).
		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
	\item {\tt paraguay}:
		Copied directly from the P2 test suite;
		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
		Written by Michael J. Sullivan.
	\item {\tt rwlock\_downgrade\_read\_test}:
		Copied directly from the P2 test suite;
		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}
		\revisionminor{(a function which atomically converts a lock held in exclusive write mode to shared read mode)}.
		Written by me (as a TA).
	\item {\tt thr\_exit\_join}:
		Copied directly from the P2 test suite;
		tests for a variety of problems between {\tt thr\_exit()} and {\tt thr\_join()},
		but especially for memory issues pertaining to stack deallocation.
\end{itemize}
\vspace{1em}

@@

For this study students were recruited remotely via email.
After each of the {\em threads} and {\em userprog} project deadlines (\sect{\ref{sec:overview-pintos}}),
\uchos staff sent students an email inviting them to volunteer to receive Landslide's bug reports,
disclaiming that it did not represent part of the official grading process but could help improve their future submissions.

@@

As described in \sect{\ref{sec:landslide-setup}},
all setup from the user's point of view is handled through the {\tt pintos-setup.sh} script.
It
and its helper {\tt pebsim/pintos/import-pintos.sh}
perform most of the same sanity checks as listed in \sect{\ref{sec:education-pebbles-instrumentation}},
then applies the patch {\tt annotate-\allowbreak{}pintos.patch}
(plus several more hacks in the script itself)
to insert the {\tt tell\_landslide()} annotations (\sect{\ref{sec:tell-landslide}})
into the student's kernel code.
The following tricks were developed after trial-and-error on student submissions from the same semester,
and serve to make sure the annotations apply consistently
to (almost) all variations of commonly-submitted code.

@@

Similar to the survey for Pebbles projects
\sect{\ref{sec:education-survey-pebbles}},
I surveyed the Pintos user study participants for their opinions.
Because of the different nature of the user study, of course,
the questions here focus more on the debugging experience than on using Landslide directly.

@@

The data set \revisionminor{comprises} Landslide's automatically-generated usage snapshots
from CMU students from semesters Spring 2015 to Spring 2018 inclusive,
% and PSU \psuos from Spring 2018, % no snapshots from psu
CMU's official project grades from same,
CMU students' survey responses and submitted project code from Fall 2017 and Spring 2018,
PSU students' survey responses and submitted project code from Spring 2018,
and U. Chicago students' survey responses from Fall 2017.
Table~\ref{tab:photo-of-ze-studence} shows the student participation rate across semesters
for the thread library study.%
\footnote{
Participation at CMU was determined by receiving any automatic usage snapshots;
participation at PSU was determined, for lack of snapshots, by sign-up form submissions, which might over-count slightly.
}
% and PSU

@@

As PSU's version of the study was planned on relatively short notice,
and without immediate access to a shared, yet confidential, academic file system (such as CMU's AFS),
I was unable to examine its students' objective usage data or bug reports for this section.
Results from PSU are presented instead in \sect{\ref{sec:education-eval-survey}}.

@@

In the recruiting lectures for every semester after the first,
I included tallies of Landslide's bug-finding achievements to date
as an extra advertisement to entice students to participate.
These included each of the measures I counted by hand as described above,
as well as aggregate totals of how many groups participated,
how many groups found bugs,
how many found at least one concurrency bug,
and how many were able to fix and subsequently verify any or all thereof.
Among the concurrency bugs, a wide variety of types were found:
deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
Table~\ref{tab:cmubugs} shows these statistics for each semester.
% may also be an overestimate if they only hacked around the bug, but that's less likely
Note that the tallies of bugs fixed may be undercounting
% lmao
%on account of the possibility
\revisionminor{because} some students may have truly fixed them but skipped the verification step thereafter.

@@

This table also shows the number of false positives, as self-reported by Landslide
as described in \sect{\ref{sec:education-pebbles-instrumentation}}.
% nah where actually doing this, where making it hapen.
%which I started counting since S'16.
Please note that this approach %to counting false positives
is limited by Landslide's ability to classify them as ``suspicious'', even if not definitely bugs.
These represent technical obstacles that either
prevented Landslide from being able to analyze the synchronization in play
(e.g., recursive {\tt mutex\_lock()} invocations)
or were deemed too pedagogically important to allow the student to proceed without fixing
(e.g., busy spin-wait synchronization loops).%
\footnote{
Landslide was also configured to issue a ``bug report''
if {\tt MAGIC\_BREAK}, a Simics debugging trap, was ever invoked,
with specific instructions to remove it;
I consider these closer in spirit to compilation errors than bug reports and so did not count them
even among false positives.
}
Other false positives arising from bugs in Landslide itself
required more individual effort to confirm;
to classify these I relied on students to help report them
during the Landslide clinics and survey,
and I report on them in \sect{\ref{sec:education-eval-survey-falsepositives}}.
Nevertheless,
apart from the spin-wait synchronization loops
(which can arise nondeterministically due to lock contention, but which Landslide already self-identifies),
I am aware of no cases of false positive {\em nondeterministic} bugs;
all unexpected false positives encountered to date were output on the first interleaving tested.
Though unscientific,
this provides some assurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
even if the deterministic tally may overcount.

@@

\begin{enumerate}
	\item {\tt priority-donate-multiple} found \revisionminor{two}
		deterministic bugs and \revisionminor{one} concurrency bug in {\tt threads} projects.
		It found the targeted priority-leak bug
		described in \sect{\ref{sec:education-pintos-tests}} in one group,
		and exposed a {\tt NULL} pointer dereference in another
		when attempting to donate priority to an already-exited thread.
		The latter bug was observed deterministically.
		A third group neglected to implement priority donation during {\tt lock\_release()} at all,
		which was found deterministically.
	\item {\tt wait-simple} found \revisionminor{one} deterministic bug
		and \revisionminor{three} concurrency bugs in {\tt userprog} projects.
		It found a deadlock in two groups' implementations,
		% for one of these groups 2 traces were emitted, same stack trace (i think) just w diff pps
		wherein {\tt process\_wait()}'s synchronization assumed
		it would always be called before {\tt process\_exit()}.
		In both cases the former's {\tt cond\_wait()} call would block forever if reordered after the latter.
		This test also found several use-after-frees for one of those two groups,
		whose {\tt process\_exit()} didn't guard against a parent process exiting and deallocating its memory first;
		this root cause manifested in heap errors in \revisionminor{three} separate locations.
		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt exec()}.
\end{enumerate}

@@

Overall, I consider all of the 7 bugs to be ``severe'':
they are all either correctness violations (priority mis-donation)
or stability issues (crash, deadlock, or data corruption).
In the upcoming survey section, although only one of these groups did the survey,
they enthusiastically reported \revisionminor{having found}
the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
Despite the small sample size, I consider this
a positive result that Landslide can be a useful concurrency programming aid
even at universities with different class projects than CMU's 15-410.
As for 1 false positive among 8 total reports,
it is difficult to decide what rate of mistakes is acceptable when the user's patience is at stake \cite{racerx},
but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
suggest that students are generally at least willing
to ask for help and make progress when technical support is available.

@@

Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
between Fall 2013 and Spring 2018,
grouped by whether or not students used Landslide during P2.
%Landslide users are, of course, the experimental group.
The control group is further distinguished by whether Landslide was offered and the student declined,
or whether the study had not yet begun that semester.
Intuition suggests students in the former \revisionminor{(``declined'')} group would be more likely
to be struggling too much to have free time to volunteer in the first place,
so any comparison between them and Landslide users is vulnerable to selection bias.
The latter control group, in which students did not have a choice, mitigates such bias,
although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
%
In addition to P2, I also show the subsequent P3 (kernel project) grades,
likewise broken down by who used Landslide previously during P2.%
\footnote{The P3 distributions are slightly smaller than those from P2,
as some students dropped the class in between.
Some new project groups also tend to appear during P3 as students either solo or find new partners,
but as these cannot be meaningfully classified as Landslide users or non-users,
I omit them here.
}
Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
one explanation could be that Landslide helped students
internalize new concurrency programming and/or debugging skills
that would help them write better kernels even without Landslide's aid.

@@

%We see in Figure~\ref{fig:photo-of-ze-studence} that, indeed,
{\bf Results.}
Overall,
Landslide users did indeed receive
slightly better grades overall on their P2 submissions than non-users from the same semesters,
although comparing to the second control group lessens that difference somewhat.
The difference is also less pronounced in both comparisons among P3 grades
(unsurprisingly, as learning lasting lessons about concurrency should be harder than fixing bugs case-by-case),
although the experimental group still maintains a tiny lead.

{\bf Statistical significance.}
Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
% TODO: run a combined A-D test for each of the two control pairs
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
I selected 4 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respectively correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
While checking for a given bug's presence by manual inspection,
I blinded myself to whether each group had been a Landslide user or not.
After unblinding, I then re-classified students who used Landslide in general,
but whose usage snapshots showed they did not run the test case in question,
as non-users.
Table~\ref{tab:eval-common-bugs} presents the results.

@@

		(see \sect{\ref{sec:education-pebbles-tests}}));

>>>>

While the previous chapter demonstrated Landslide's bug-finding power
compared to prior MC techniques in a controlled environment,
whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
And while my M.S. thesis~\cite{landslide} showed that students
could annotate P3 Pebbles kernels and thence use Landslide to debug them,
the annotations alone required 2 hours of effort on average per user,
meaning the only students who could benefit were the ones already succeeding enough to have such free time.
Since then, I have extended Landslide with a fully-automatic instrumentation process
for Pebbles thread libraries (P2s) (\cref{sec:education-pebbles-instrumentation})
and Pintos kernels (\cref{sec:education-pintos-instrumentation})
to improve its accessibility.

I have run several user studies in the Operating Systems classes
at Carnegie Mellon University (CMU), University of Chicago (U. Chicago),
and \revisionminor{The Pennsylvania} State University (PSU),
wherein students get to use Landslide to find and diagnose their own bugs during the semester.
At CMU, I analyzed logs and code snapshots taken as students used Landslide during P2
(\cref{sec:education-eval-bugfinding}),
as well as the grades ultimately assigned after students who either did or didn't use Landslide submitted their projects
(\cref{sec:education-eval-grades}).
At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s
(\cref{sec:education-eval-survey}).
At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
then \revisionminor{they} returned any resulting bug reports to students
(\cref{sec:education-eval-bugfinding})
and likewise surveyed them on the quality of the diagnostic output
(\cref{sec:education-eval-survey}).

@@

As described in \cref{sec:landslide-setup},
all setup from the user's point of view is handled through the {\tt p2-setup.sh} script.%
\footnote{PSU's version is called {\tt psu-setup.sh};
in this section {\tt p2-setup.sh} refers to both unless otherwise noted.
}
It, its helper scripts (\cref{sec:landslide-glue}),
and the {\tt landslide} script itself contain several checks to prevent
studence
from accidentally misusing Landslide in ways that could produce mysterious crashes, false bug reports, and so on
(the need for each \revisionminor{check}, as the reader might imagine, discovered through bitter experience).
These include:
\begin{itemize}
	\item {\tt p2-setup.sh} checks if the directory argument correctly points at the top-level P2 basecode directory
		rather than any subdirectories such as {\tt user/libthread/}.
	\item {\tt check-need-p2-setup-again.sh} checks if any source files in the original P2 source directory
		(the argument supplied to {\tt p2-setup.sh}) \revisionminor{have been updated},
		in case the student hoped to fix some bug and verify their fix but forgot to re-run the setup script.
	\item {\tt landslide} checks the supplied test name matches one of the endorsed Landslide-friendly tests
		(students love trying to run Landslide with {\tt racer}, {\tt largetest},
		or even the string {\tt OPTIONS}).
	\item {\tt landslide} checks if any other instance of itself is simultaneously running in the same directory,
		and if so, refuses to do so and advises the student
		to {\tt git clone} the repository afresh for simultaneous use.%
		\footnote{This is ironically implemented with a non-atomic lock file
		and should really be using {\tt flock} instead.
		}
\end{itemize}
\vspace{1em}

\noindent Landslide also includes several P2-specific instrumentations and features to cope with various student irregularities:
\begin{itemize}
	\item Quicksand emits different combinations of {\tt within\_function}/{\tt without\_function} directives
		for Landslide depending on the name of the test.
		For example, for {\tt paradise\_\allowbreak{}lost},
		\revisionminor{designed to test student semaphore code,}
		Landslide will not preempt in a function named {\tt critical\_section()},
		which the test case uses to protect an internal counter used to detect the bug;
		and it will not preempt in any of the {\tt thr\_*()} thread library API functions
		for tests intended to target just the concurrency primitives.
		In future work this could be improved as annotations to be placed inside the test case code itself.
	\item Landslide finds ad-hoc synchronization patterns,
		such as {\tt while (!flag) yield()} or {\tt while (xchg(...)) continue},
		which students often open-code rather than using the prescribed synchronization API,
		and treats them as synchronization points as described in \cref{sec:landslide-blocking}.
	\item Landslide finds ``too suspicious'' spinwait-loops in mutex implementations
		which are neither yield- nor {\tt xchg}-loops (as described above),
		which would ordinarily be classified as infinite loop bugs,
		and reports them with a suggestive message ({\tt undesirable\_\allowbreak{}loop\_html()} in {\tt landslide.c})
		referring the student to the appropriate lecture material
		\cite{synchronization-2}.
	\item The {\tt landslide} wrapper script logs the time and command-line options of invocation
		and captures a snapshot of the student code and results of the test and saves them to AFS
		\revisionminor{(CMU's network file system)~\cite{afs}}
		after each run.
		% HURDLE_VIOLATION -- no, this is for p3 (context switcher) features only
\end{itemize}

@@

Landslide ships with several ``approved'' test cases,
i.e., programs copied from, derived from, or at least vaguely resembling
the tests distributed with P2,
which I curated to produce concurrent behaviour suitable for stateless model checking.
Some tests are crafted to target specific bugs which,
from personal experience as a TA, are common in many student submissions;
others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
Many use some of the features/annotations described in \cref{sec:landslide-testcases}.

@@

The following tests were released to CMU students:
\begin{itemize}
	\item {\tt broadcast\_test}:
		Tests the {\tt cond\_broadcast()} signalling path with a single waiter.
	\item {\tt mutex\_test}:
		Tests student mutexes under 2 threads with 2 iterations
		(the \revisionminor{second} iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
		This test uses the {\tt TESTING\_MUTEXES}
		described in \cref{sec:landslide-staticconfig}
		to enable data-race preemption points within the mutex implementation.
	\item {\tt paradise\_lost}:
		Written for the sake of the Landslide lecture demo
		(\cref{sec:education-pebbles-recruiting}).
		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
	\item {\tt paraguay}:
		Copied directly from the P2 test suite;
		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
		Written by Michael J. Sullivan.
	\item {\tt rwlock\_downgrade\_read\_test}:
		Copied directly from the P2 test suite;
		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}
		\revisionminor{(a function which atomically converts a lock held in exclusive write mode to shared read mode)}.
		Written by me (as a TA).
	\item {\tt thr\_exit\_join}:
		Copied directly from the P2 test suite;
		tests for a variety of problems between {\tt thr\_exit()} and {\tt thr\_join()},
		but especially for memory issues pertaining to stack deallocation.
\end{itemize}
\vspace{1em}

@@

For this study students were recruited remotely via email.
After each of the {\em threads} and {\em userprog} project deadlines (\cref{sec:overview-pintos}),
\uchos staff sent students an email inviting them to volunteer to receive Landslide's bug reports,
disclaiming that it did not represent part of the official grading process but could help improve their future submissions.

@@

As described in \cref{sec:landslide-setup},
all setup from the user's point of view is handled through the {\tt pintos-setup.sh} script.
It
and its helper {\tt pebsim/pintos/import-pintos.sh}
perform most of the same sanity checks as listed in \cref{sec:education-pebbles-instrumentation},
then applies the patch {\tt annotate-\allowbreak{}pintos.patch}
(plus several more hacks in the script itself)
to insert the {\tt tell\_landslide()} annotations (\cref{sec:tell-landslide})
into the student's kernel code.
The following tricks were developed after trial-and-error on student submissions from the same semester,
and serve to make sure the annotations apply consistently
to (almost) all variations of commonly-submitted code.

@@

Similar to the survey for Pebbles projects
\cref{sec:education-survey-pebbles},
I surveyed the Pintos user study participants for their opinions.
Because of the different nature of the user study, of course,
the questions here focus more on the debugging experience than on using Landslide directly.

@@

The data set \revisionminor{comprises} Landslide's automatically-generated usage snapshots
from CMU students from semesters Spring 2015 to Spring 2018 inclusive,
% and PSU \psuos from Spring 2018, % no snapshots from psu
CMU's official project grades from same,
CMU students' survey responses and submitted project code from Fall 2017 and Spring 2018,
PSU students' survey responses and submitted project code from Spring 2018,
and U. Chicago students' survey responses from Fall 2017.
\Cref{tab:photo-of-ze-studence} shows the student participation rate across semesters
for the thread library study.%
\footnote{
Participation at CMU was determined by receiving any automatic usage snapshots;
participation at PSU was determined, for lack of snapshots, by sign-up form submissions, which might over-count slightly.
}
% and PSU

@@

As PSU's version of the study was planned on relatively short notice,
and without immediate access to a shared, yet confidential, academic file system (such as CMU's AFS),
I was unable to examine its students' objective usage data or bug reports for this section.
Results from PSU are presented instead in \cref{sec:education-eval-survey}.

@@

In the recruiting lectures for every semester after the first,
I included tallies of Landslide's bug-finding achievements to date
as an extra advertisement to entice students to participate.
These included each of the measures I counted by hand as described above,
as well as aggregate totals of how many groups participated,
how many groups found bugs,
how many found at least one concurrency bug,
and how many were able to fix and subsequently verify any or all thereof.
Among the concurrency bugs, a wide variety of types were found:
deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
\Cref{tab:cmubugs} shows these statistics for each semester.
% may also be an overestimate if they only hacked around the bug, but that's less likely
Note that the tallies of bugs fixed may be undercounting
% lmao
%on account of the possibility
\revisionminor{because} some students may have truly fixed them but skipped the verification step thereafter.

@@

This table also shows the number of false positives, as self-reported by Landslide
as described in \cref{sec:education-pebbles-instrumentation}.
% nah where actually doing this, where making it hapen.
%which I started counting since S'16.
Please note that this approach %to counting false positives
is limited by Landslide's ability to classify them as ``suspicious'', even if not definitely bugs.
These represent technical obstacles that either
prevented Landslide from being able to analyze the synchronization in play
(e.g., recursive {\tt mutex\_lock()} invocations)
or were deemed too pedagogically important to allow the student to proceed without fixing
(e.g., busy spin-wait synchronization loops).%
\footnote{
Landslide was also configured to issue a ``bug report''
if {\tt MAGIC\_BREAK}, a Simics debugging trap, was ever invoked,
with specific instructions to remove it;
I consider these closer in spirit to compilation errors than bug reports and so did not count them
even among false positives.
}
Other false positives arising from bugs in Landslide itself
required more individual effort to confirm;
to classify these I relied on students to help report them
during the Landslide clinics and survey,
and I report on them in \cref{sec:education-eval-survey-falsepositives}.
Nevertheless,
apart from the spin-wait synchronization loops
(which can arise nondeterministically due to lock contention, but which Landslide already self-identifies),
I am aware of no cases of false positive {\em nondeterministic} bugs;
all unexpected false positives encountered to date were output on the first interleaving tested.
Though unscientific,
this provides some assurance that \Cref{tab:cmubugs}'s count of concurrency bugs is accurate,
even if the deterministic tally may overcount.

@@

\begin{enumerate}
	\item {\tt priority-donate-multiple} found \revisionminor{two}
		deterministic bugs and \revisionminor{one} concurrency bug in {\tt threads} projects.
		It found the targeted priority-leak bug
		described in \cref{sec:education-pintos-tests} in one group,
		and exposed a {\tt NULL} pointer dereference in another
		when attempting to donate priority to an already-exited thread.
		The latter bug was observed deterministically.
		A third group neglected to implement priority donation during {\tt lock\_release()} at all,
		which was found deterministically.
	\item {\tt wait-simple} found \revisionminor{one} deterministic bug
		and \revisionminor{three} concurrency bugs in {\tt userprog} projects.
		It found a deadlock in two groups' implementations,
		% for one of these groups 2 traces were emitted, same stack trace (i think) just w diff pps
		wherein {\tt process\_\allowbreak{}wait()}'s synchronization assumed
		it would always be called before {\tt process\_exit()}.
		In both cases the former's {\tt cond\_wait()} call would block forever if reordered after the latter.
		This test also found several use-after-frees for one of those two groups,
		whose {\tt process\_exit()} didn't guard against a parent process exiting and deallocating its memory first;
		this root cause manifested in heap errors in \revisionminor{three} separate locations.
		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt exec()}.
\end{enumerate}

@@

Overall, I consider all of the 7 bugs to be ``severe'':
they are all either correctness violations (priority mis-donation)
or stability issues (crash, deadlock, or data corruption).
In the upcoming survey section, although only one of these groups did the survey,
they enthusiastically reported \revisionminor{having found}
the bug reports very helpful (\cref{sec:education-survey-response-data}).
Despite the small sample size, I consider this
a positive result that Landslide can be a useful concurrency programming aid
even at universities with different class projects than CMU's 15-410.
As for 1 false positive among 8 total reports,
it is difficult to decide what rate of mistakes is acceptable when the user's patience is at stake \cite{racerx},
but the upcoming survey results (\cref{sec:education-eval-survey-falsepositives})
suggest that students are generally at least willing
to ask for help and make progress when technical support is available.

@@

\Cref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
between Fall 2013 and Spring 2018,
grouped by whether or not students used Landslide during P2.
%Landslide users are, of course, the experimental group.
The control group is further distinguished by whether Landslide was offered and the student declined,
or whether the study had not yet begun that semester.
Intuition suggests students in the former \revisionminor{(``declined'')} group would be more likely
to be struggling too much to have free time to volunteer in the first place,
so any comparison between them and Landslide users is vulnerable to selection bias.
The latter control group, in which students did not have a choice, mitigates such bias,
although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
%
In addition to P2, I also show the subsequent P3 (kernel project) grades,
likewise broken down by who used Landslide previously during P2.%
\footnote{The P3 distributions are slightly smaller than those from P2,
as some students dropped the class in between.
Some new project groups also tend to appear during P3 as students either solo or find new partners,
but as these cannot be meaningfully classified as Landslide users or non-users,
I omit them here.
}
Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
one explanation could be that Landslide helped students
internalize new concurrency programming and/or debugging skills
that would help them write better kernels even without Landslide's aid.

@@

%We see in \Cref{fig:photo-of-ze-studence} that, indeed,
{\bf Results.}
Overall,
Landslide users did indeed receive
slightly better grades overall on their P2 submissions than non-users from the same semesters,
although comparing to the second control group lessens that difference somewhat.
The difference is also less pronounced in both comparisons among P3 grades
(unsurprisingly, as learning lasting lessons about concurrency should be harder than fixing bugs case-by-case),
although the experimental group still maintains a tiny lead.

{\bf Statistical significance.}
\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
% TODO: run a combined A-D test for each of the two control pairs
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

Based on the results from \cref{sec:education-eval-bugfinding},
I selected 4 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respectively correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
While checking for a given bug's presence by manual inspection,
I blinded myself to whether each group had been a Landslide user or not.
After unblinding, I then re-classified students who used Landslide in general,
but whose usage snapshots showed they did not run the test case in question,
as non-users.
\Cref{tab:eval-common-bugs} presents the results.

@@

		(see \cref{sec:education-pebbles-tests}));
